---
jupyter:
  jupytext:
    formats: ipynb,Rmd
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.1'
      jupytext_version: 1.1.1
  kernelspec:
    display_name: R
    language: R
    name: ir
---

```{r}
# Libraries
suppressPackageStartupMessages(library(tidyverse))

# Global options
options(warn = -1)

# Loading the data
load("DF.Rdata")
```

# Clinicopathologic and outcome features

```{r}
# Glimpse of the clinicopathologic dataset
glimpse(DF_Clinical)
```

```{r}
# Functions for describing clinicopathologic features
# Argument 'x' refers to the variable under evaluation

# Summary for a numeric variable
summary_num_x <- function(data, x) {

    x <- enquo(x)
    
    num_x <- data %>%
        summarize(
            N = n(),
            Mean = mean(!! x, na.rm = TRUE),
            SD = sd(!! x, na.rm = TRUE),
            Median = median(!! x, na.rm = TRUE),
            IQR = IQR(!! x, na.rm = TRUE),
            Min = min(!! x, na.rm = TRUE),
            Max = max(!! x, na.rm = TRUE),
            Missing = sum(is.na(!! x))
        )
    
    print(num_x)
}

# Summary for a factor (categorical) variable
summary_fct_x <- function(data, x) {
    
    x <- enquo(x)
    
    fct_x <- data %>%
        count(!! x) %>%
        mutate(freq = n / sum(n) * 100) %>%
        rename(Levels = !! x) %>%
        mutate(freq = round(freq,  digits = 1))

    print(fct_x)

}
```

## Clinical features
### Age, in years
```{r}
DF_Clinical %>% summary_num_x(age)
```

### Sex
```{r}
DF_Clinical %>% summary_fct_x(sex)
```

### Race
```{r}
DF_Clinical %>% summary_fct_x(race)
```

## Pathologic features
### Fuhrman grade
```{r}
DF_Clinical %>% summary_fct_x(fuhrman)
```

### pT stage
```{r}
DF_Clinical %>% summary_fct_x(pt_stage)
```


### Tumor size, in cm
```{r}
DF_Clinical %>% summary_num_x(size)
```

### Tumor focality
```{r}
DF_Clinical %>% summary_fct_x(focality)
```

## Outcome features
### Follow-up, in months
```{r}
DF_Clinical %>% summary_num_x(follow_up)
```

### Tumor progression
```{r}
DF_Clinical %>% summary_fct_x(progression)
```

### Overall mortality
```{r}
DF_Clinical %>% summary_fct_x(death)
```

### Cancer-related mortality
```{r}
DF_Clinical %>% summary_fct_x(dod)
```

# Expression by tissue type
```{r}
glimpse(DF)
```

```{r}
# A function to summarize biomarkers expression by tissue type and test the associations
# 'x' refers to the expression in nontumor tissues
# 'y' refers to the expression in primary tumors
# 'z' refers to the expression in metastatic tumors

summary_by <- function(data, x, y, z) {
    
    x <- enquo(x)
    y <- enquo(y)
    z <- enquo(z)
    
    # Tidying the data
    df <- data %>%
        select(!! x, !! y, !! z) %>%
        gather(tissue_type, biomarker_expression) %>%
        mutate(
            tissue_type = str_replace_all(
                tissue_type, c(
                    ".*_normal.*" = "Non-tumor epithelium",
                    ".*_tumor.*" = "Primary tumor",
                    ".*_metastatic.*" = "Metastatic tumor"
                )
            ),
            tissue_type = fct_relevel(
                tissue_type,
                "Metastatic tumor", after = Inf
            )
        )
    
    # Creating summary table
    summary_biomarker <- df %>%
        group_by(tissue_type) %>%
        summarize(
            N = n(),
            Mean = mean(biomarker_expression, na.rm = TRUE),
            SD = sd(biomarker_expression, na.rm = TRUE),
            Median = median(biomarker_expression, na.rm = TRUE),
            IQR = IQR(biomarker_expression, na.rm = TRUE),
            Min = min(biomarker_expression, na.rm = TRUE),
            Max = max(biomarker_expression, na.rm = TRUE),
            Missing = sum(is.na(biomarker_expression))
        ) %>%
        rename(
            "Tissue type" = tissue_type
        )

    # Testing for overall associations
    kw_biomarker <- kruskal.test(
        formula = biomarker_expression ~ tissue_type,
        data = df
    )
    
    # Testing for pairwise associations
    pairwise_biomarker <- with(
        df,
        pairwise.wilcox.test(
            x = biomarker_expression,
            g = tissue_type,
            p.adjust.method = "bonferroni"
        )
    )
    
    # Printing the results
    print(summary_biomarker)
    print(kw_biomarker)
    print(pairwise_biomarker)
    
}
```

## PD-L1
### Epithelial cells
```{r}
DF %>% summary_by(pdl1_normal, pdl1_tumor, pdl1_metastatic)
```

### Peritumoral lymphocytes
```{r}
DF %>% summary_by(pdl1_normal_stroma, pdl1_tumor_stroma, pdl1_metastatic_stroma)
```

## FOXP3
### Intratumoral lymphocytes
```{r}
DF %>% summary_by(foxp3_normal_lymph, foxp3_tumor_lymph, foxp3_metastatic_lymph)
```

### Peritumoral lymphocytes
```{r}
DF %>% summary_by(foxp3_normal_lymph_stroma, foxp3_tumor_lymph_stroma, foxp3_metastatic_lymph_stroma)
```

### Tumor cells
```{r}
DF %>% summary_by(foxp3_normal, foxp3_tumor, foxp3_metastatic)
```

## CD8
### Intratumoral lymphocytes
```{r}
DF %>% summary_by(cd8_normal, cd8_tumor, cd8_metastatic)
```


### Peritumoral lymphocytes
```{r}
DF %>% summary_by(cd8_normal_stroma, cd8_tumor_stroma, cd8_metastatic_stroma)
```

## CD8/Ki67
### Intratumoral lymphocytes
```{r}
DF %>% summary_by(cd8ki67_normal, cd8ki67_tumor, cd8ki67_metastatic)
```

### Peritumoral lymphocytes
```{r}
DF %>% summary_by(cd8ki67_normal_stroma, cd8ki67_tumor_stroma, cd8ki67_metastatic_stroma)
```

## Ki67
### Intratumoral lymphocytes
```{r}
DF %>% summary_by(ki67_normal, ki67_tumor, ki67_metastatic)
```

# Expression by clinicopathologic features
```{r}
# A function to define the summarizing statistic for biomarkers
summarize_biomarker <- function(x) { quantile(x, probs = 1, na.rm = TRUE) }

# Summaries by maximum tumor expression
DF_ASSOCIATION <- DF %>%
    group_by(sp) %>%
    summarize(
        age_group = unique(age_group),
        sex = unique(sex),
        race = unique(race),
        fuhrman = unique(fuhrman),
        pt_stage = unique(pt_stage),
        size_group = unique(size_group),
        focality = unique(focality),
        progression = unique(progression),
        death = unique(death),
        dod = unique(dod),
        pdl1_tumor = summarize_biomarker(pdl1_tumor),
        pdl1_intratumoral = summarize_biomarker(pdl1_tumor_stroma),
        foxp3_intratumoral = summarize_biomarker(foxp3_tumor_lymph),
        foxp3_peritumoral = summarize_biomarker(foxp3_tumor_lymph_stroma),
        cd8_intratumoral = summarize_biomarker(cd8_tumor),
        cd8_peritumoral = summarize_biomarker(cd8_tumor_stroma),
        cd8ki67_intratumoral = summarize_biomarker(cd8ki67_tumor),
        cd8ki67_peritumoral = summarize_biomarker(cd8ki67_tumor_stroma),
        ki67_intratumoral = summarize_biomarker(ki67_tumor)
    )

glimpse(DF_ASSOCIATION)
```

```{r}
# A function to evaluate the association between biomarkers expression and clinicopathologic features
summary_association <- function(data, biomarker, feature)  {
    
    biomarker <- enquo(biomarker)
    feature <- enquo(feature)
    df <- data %>%
        mutate(
            biomarker = !! biomarker,
            feature = !! feature
        )
    
    # Creating summary table
    tbl <- df %>%
        group_by(feature) %>%
        summarize(
            N = n(),
            Mean = mean(biomarker, na.rm = TRUE),
            SD = sd(biomarker, na.rm = TRUE),
            Median = median(biomarker, na.rm = TRUE),
            IQR = IQR(biomarker, na.rm = TRUE),
            Min = min(biomarker, na.rm = TRUE),
            Max = max(biomarker, na.rm = TRUE),
            Missing = sum(is.na(biomarker))
        )
    
    # Testing for association
    if (nrow(tbl) == 2) {
        test <- wilcox.test(biomarker ~ feature, df)
    }
    if (nrow(tbl) > 2) {
        test <- kruskal.test(biomarker ~ feature, df)
    }
    
    # Testing for pairwise associations
    if (nrow(tbl) > 2) {
        test_pairwise <- with(df, pairwise.wilcox.test(biomarker, feature, p.adjust.method = "bonferroni"))
    }
    
    # Printing the results
    print(tbl)
    print(test)
    if (nrow(tbl) > 2) { print(test_pairwise) }
    
}
```

## PD-L1
### Tumor cells
```{r}
DF_ASSOCIATION <- DF_ASSOCIATION %>% 
  mutate(marker = pdl1_tumor)
```

#### Age group, median
```{r}
DF_ASSOCIATION %>% summary_association(marker, age_group)
```

#### Sex
```{r}
DF_ASSOCIATION %>% summary_association(marker, sex)
```

#### Race
```{r}
DF_ASSOCIATION %>% filter(!is.na(race)) %>% summary_association(marker, race)
```

#### Fuhrman grade
```{r}
DF_ASSOCIATION %>% filter(!is.na(fuhrman)) %>% summary_association(marker, fuhrman)
```

#### pT stage
```{r}
DF_ASSOCIATION %>% filter(!is.na(pt_stage)) %>% summary_association(marker, pt_stage)
```

#### Tumor size, median
```{r}
DF_ASSOCIATION %>% filter(!is.na(size_group)) %>% summary_association(marker, size_group)
```

#### Tumor focality
```{r}
DF_ASSOCIATION %>% filter(!is.na(focality)) %>% summary_association(marker, focality)
```

#### Tumor progression
```{r}
DF_ASSOCIATION %>% filter(!is.na(progression)) %>% summary_association(marker, progression)
```

#### Overall mortality
```{r}
DF_ASSOCIATION %>% filter(!is.na(death)) %>% summary_association(marker, death)
```

#### Cancer-related mortality
```{r}
DF_ASSOCIATION %>% filter(!is.na(dod)) %>% summary_association(marker, dod)
```

### Intratumoral lymphocytes
```{r}
DF_ASSOCIATION <- DF_ASSOCIATION %>% 
  mutate(marker = pdl1_intratumoral)
```

#### Age group, median
```{r}
DF_ASSOCIATION %>% summary_association(marker, age_group)
```

#### Sex
```{r}
DF_ASSOCIATION %>% summary_association(marker, sex)
```

#### Race
```{r}
DF_ASSOCIATION %>% filter(!is.na(race)) %>% summary_association(marker, race)
```

#### Fuhrman grade
```{r}
DF_ASSOCIATION %>% filter(!is.na(fuhrman)) %>% summary_association(marker, fuhrman)
```

#### pT stage
```{r}
DF_ASSOCIATION %>% filter(!is.na(pt_stage)) %>% summary_association(marker, pt_stage)
```

#### Tumor size, median
```{r}
DF_ASSOCIATION %>% filter(!is.na(size_group)) %>% summary_association(marker, size_group)
```

#### Tumor focality
```{r}
DF_ASSOCIATION %>% filter(!is.na(focality)) %>% summary_association(marker, focality)
```

#### Tumor progression
```{r}
DF_ASSOCIATION %>% filter(!is.na(progression)) %>% summary_association(marker, progression)
```

#### Overall mortality
```{r}
DF_ASSOCIATION %>% filter(!is.na(death)) %>% summary_association(marker, death)
```

#### Cancer-related mortality
```{r}
DF_ASSOCIATION %>% filter(!is.na(dod)) %>% summary_association(marker, dod)
```

## FOXP3
### Intratumoral lymphocytes
```{r}
DF_ASSOCIATION <- DF_ASSOCIATION %>% 
  mutate(marker = foxp3_intratumoral)
```

#### Age group, median
```{r}
DF_ASSOCIATION %>% summary_association(marker, age_group)
```

#### Sex
```{r}
DF_ASSOCIATION %>% summary_association(marker, sex)
```

#### Race
```{r}
DF_ASSOCIATION %>% filter(!is.na(race)) %>% summary_association(marker, race)
```

#### Fuhrman grade
```{r}
DF_ASSOCIATION %>% filter(!is.na(fuhrman)) %>% summary_association(marker, fuhrman)
```

#### pT stage
```{r}
DF_ASSOCIATION %>% filter(!is.na(pt_stage)) %>% summary_association(marker, pt_stage)
```

#### Tumor size, median
```{r}
DF_ASSOCIATION %>% filter(!is.na(size_group)) %>% summary_association(marker, size_group)
```

#### Tumor focality
```{r}
DF_ASSOCIATION %>% filter(!is.na(focality)) %>% summary_association(marker, focality)
```

#### Tumor progression
```{r}
DF_ASSOCIATION %>% filter(!is.na(progression)) %>% summary_association(marker, progression)
```

#### Overall mortality
```{r}
DF_ASSOCIATION %>% filter(!is.na(death)) %>% summary_association(marker, death)
```

#### Cancer-related mortality
```{r}
DF_ASSOCIATION %>% filter(!is.na(dod)) %>% summary_association(marker, dod)
```

### Peritumoral lymphocytes
```{r}
DF_ASSOCIATION <- DF_ASSOCIATION %>% 
  mutate(marker = foxp3_peritumoral)
```

#### Age group, median
```{r}
DF_ASSOCIATION %>% summary_association(marker, age_group)
```

#### Sex
```{r}
DF_ASSOCIATION %>% summary_association(marker, sex)
```

#### Race
```{r}
DF_ASSOCIATION %>% filter(!is.na(race)) %>% summary_association(marker, race)
```

#### Fuhrman grade
```{r}
DF_ASSOCIATION %>% filter(!is.na(fuhrman)) %>% summary_association(marker, fuhrman)
```

#### pT stage
```{r}
DF_ASSOCIATION %>% filter(!is.na(pt_stage)) %>% summary_association(marker, pt_stage)
```

#### Tumor size, median
```{r}
DF_ASSOCIATION %>% filter(!is.na(size_group)) %>% summary_association(marker, size_group)
```

#### Tumor focality
```{r}
DF_ASSOCIATION %>% filter(!is.na(focality)) %>% summary_association(marker, focality)
```

#### Tumor progression
```{r}
DF_ASSOCIATION %>% filter(!is.na(progression)) %>% summary_association(marker, progression)
```

#### Overall mortality
```{r}
DF_ASSOCIATION %>% filter(!is.na(death)) %>% summary_association(marker, death)
```

#### Cancer-related mortality
```{r}
DF_ASSOCIATION %>% filter(!is.na(dod)) %>% summary_association(marker, dod)
```

## CD8
### Intratumoral lymphocytes
```{r}
DF_ASSOCIATION <- DF_ASSOCIATION %>% 
  mutate(marker = cd8_intratumoral)
```

#### Age group, median
```{r}
DF_ASSOCIATION %>% summary_association(marker, age_group)
```

#### Sex
```{r}
DF_ASSOCIATION %>% summary_association(marker, sex)
```

#### Race
```{r}
DF_ASSOCIATION %>% filter(!is.na(race)) %>% summary_association(marker, race)
```

#### Fuhrman grade
```{r}
DF_ASSOCIATION %>% filter(!is.na(fuhrman)) %>% summary_association(marker, fuhrman)
```

#### pT stage
```{r}
DF_ASSOCIATION %>% filter(!is.na(pt_stage)) %>% summary_association(marker, pt_stage)
```

#### Tumor size, median
```{r}
DF_ASSOCIATION %>% filter(!is.na(size_group)) %>% summary_association(marker, size_group)
```

#### Tumor focality
```{r}
DF_ASSOCIATION %>% filter(!is.na(focality)) %>% summary_association(marker, focality)
```

#### Tumor progression
```{r}
DF_ASSOCIATION %>% filter(!is.na(progression)) %>% summary_association(marker, progression)
```

#### Overall mortality
```{r}
DF_ASSOCIATION %>% filter(!is.na(death)) %>% summary_association(marker, death)
```

#### Cancer-related mortality
```{r}
DF_ASSOCIATION %>% filter(!is.na(dod)) %>% summary_association(marker, dod)
```

### Peritumoral lymphocytes
```{r}
DF_ASSOCIATION <- DF_ASSOCIATION %>% 
  mutate(marker = cd8_peritumoral)
```

#### Age group, median
```{r}
DF_ASSOCIATION %>% summary_association(marker, age_group)
```

#### Sex
```{r}
DF_ASSOCIATION %>% summary_association(marker, sex)
```

#### Race
```{r}
DF_ASSOCIATION %>% filter(!is.na(race)) %>% summary_association(marker, race)
```

#### Fuhrman grade
```{r}
DF_ASSOCIATION %>% filter(!is.na(fuhrman)) %>% summary_association(marker, fuhrman)
```

#### pT stage
```{r}
DF_ASSOCIATION %>% filter(!is.na(pt_stage)) %>% summary_association(marker, pt_stage)
```

#### Tumor size, median
```{r}
DF_ASSOCIATION %>% filter(!is.na(size_group)) %>% summary_association(marker, size_group)
```

#### Tumor focality
```{r}
DF_ASSOCIATION %>% filter(!is.na(focality)) %>% summary_association(marker, focality)
```

#### Tumor progression
```{r}
DF_ASSOCIATION %>% filter(!is.na(progression)) %>% summary_association(marker, progression)
```

#### Overall mortality
```{r}
DF_ASSOCIATION %>% filter(!is.na(death)) %>% summary_association(marker, death)
```

#### Cancer-related mortality
```{r}
DF_ASSOCIATION %>% filter(!is.na(dod)) %>% summary_association(marker, dod)
```

## CD8/Ki67
### Intratumoral lymphocytes
```{r}
DF_ASSOCIATION <- DF_ASSOCIATION %>% 
  mutate(marker = cd8ki67_intratumoral)
```

#### Age group, median
```{r}
DF_ASSOCIATION %>% summary_association(marker, age_group)
```

#### Sex
```{r}
DF_ASSOCIATION %>% summary_association(marker, sex)
```

#### Race
```{r}
DF_ASSOCIATION %>% filter(!is.na(race)) %>% summary_association(marker, race)
```

#### Fuhrman grade
```{r}
DF_ASSOCIATION %>% filter(!is.na(fuhrman)) %>% summary_association(marker, fuhrman)
```

#### pT stage
```{r}
DF_ASSOCIATION %>% filter(!is.na(pt_stage)) %>% summary_association(marker, pt_stage)
```

#### Tumor size, median
```{r}
DF_ASSOCIATION %>% filter(!is.na(size_group)) %>% summary_association(marker, size_group)
```

#### Tumor focality
```{r}
DF_ASSOCIATION %>% filter(!is.na(focality)) %>% summary_association(marker, focality)
```

#### Tumor progression
```{r}
DF_ASSOCIATION %>% filter(!is.na(progression)) %>% summary_association(marker, progression)
```

#### Overall mortality
```{r}
DF_ASSOCIATION %>% filter(!is.na(death)) %>% summary_association(marker, death)
```

#### Cancer-related mortality
```{r}
DF_ASSOCIATION %>% filter(!is.na(dod)) %>% summary_association(marker, dod)
```

### Peritumoral lymphocytes
```{r}
DF_ASSOCIATION <- DF_ASSOCIATION %>% 
  mutate(marker = cd8ki67_peritumoral)
```

#### Age group, median
```{r}
DF_ASSOCIATION %>% summary_association(marker, age_group)
```

#### Sex
```{r}
DF_ASSOCIATION %>% summary_association(marker, sex)
```

#### Race
```{r}
DF_ASSOCIATION %>% filter(!is.na(race)) %>% summary_association(marker, race)
```

#### Fuhrman grade
```{r}
DF_ASSOCIATION %>% filter(!is.na(fuhrman)) %>% summary_association(marker, fuhrman)
```

#### pT stage
```{r}
DF_ASSOCIATION %>% filter(!is.na(pt_stage)) %>% summary_association(marker, pt_stage)
```

#### Tumor size, median
```{r}
DF_ASSOCIATION %>% filter(!is.na(size_group)) %>% summary_association(marker, size_group)
```

#### Tumor focality
```{r}
DF_ASSOCIATION %>% filter(!is.na(focality)) %>% summary_association(marker, focality)
```

#### Tumor progression
```{r}
DF_ASSOCIATION %>% filter(!is.na(progression)) %>% summary_association(marker, progression)
```

#### Overall mortality
```{r}
DF_ASSOCIATION %>% filter(!is.na(death)) %>% summary_association(marker, death)
```

#### Cancer-related mortality
```{r}
DF_ASSOCIATION %>% filter(!is.na(dod)) %>% summary_association(marker, dod)
```

## Ki67
### Intratumoral lymphocytes
```{r}
DF_ASSOCIATION <- DF_ASSOCIATION %>% 
  mutate(marker = ki67_intratumoral)
```

#### Age group, median
```{r}
DF_ASSOCIATION %>% summary_association(marker, age_group)
```

#### Sex
```{r}
DF_ASSOCIATION %>% summary_association(marker, sex)
```

#### Race
```{r}
DF_ASSOCIATION %>% filter(!is.na(race)) %>% summary_association(marker, race)
```

#### Fuhrman grade
```{r}
DF_ASSOCIATION %>% filter(!is.na(fuhrman)) %>% summary_association(marker, fuhrman)
```

#### pT stage
```{r}
DF_ASSOCIATION %>% filter(!is.na(pt_stage)) %>% summary_association(marker, pt_stage)
```

#### Tumor size, median
```{r}
DF_ASSOCIATION %>% filter(!is.na(size_group)) %>% summary_association(marker, size_group)
```

#### Tumor focality
```{r}
DF_ASSOCIATION %>% filter(!is.na(focality)) %>% summary_association(marker, focality)
```

#### Tumor progression
```{r}
DF_ASSOCIATION %>% filter(!is.na(progression)) %>% summary_association(marker, progression)
```

#### Overall mortality
```{r}
DF_ASSOCIATION %>% filter(!is.na(death)) %>% summary_association(marker, death)
```

#### Cancer-related mortality
```{r}
DF_ASSOCIATION %>% filter(!is.na(dod)) %>% summary_association(marker, dod)
```
