---
jupyter:
  jupytext:
    formats: ipynb,Rmd
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.1'
      jupytext_version: 1.1.1
  kernelspec:
    display_name: R
    language: R
    name: ir
---

```{r}
# Libraries
suppressPackageStartupMessages(library(tidyverse))

# Options
options(warn = -1)

# Scripts
source("summarize.R")

# Data
load("DF.Rdata")

OUTCOME <- DF %>% 
  select(
    age_group,
    stage_group,
    fuhrman_group,
    death:progression,
    cd8foxp3_intratumoral:ki67_intratumoral_level
  )

glimpse(OUTCOME)
```

# PD-L1 expression
## Tumor cells
```{r}
OUTCOME <- OUTCOME %>% 
  mutate(marker = pdl1_tumor_level)
```

### Tumor progression
#### Unadjusted model
```{r}
OUTCOME %>% OR(progression ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(progression ~ marker + age_group + stage_group + fuhrman_group)
```

### Overall mortality
#### Unadjusted model
```{r}
OUTCOME %>% OR(death ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(death ~ marker + age_group + stage_group + fuhrman_group)
```

### Cancer-related mortality
#### Unadjusted model
```{r}
OUTCOME %>% OR(dod ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(dod ~ marker + age_group + stage_group + fuhrman_group)
```

## Intratumoral lymphocytes
```{r}
OUTCOME <- OUTCOME %>% 
  mutate(marker = pdl1_intratumoral_level)
```

### Tumor progression
#### Unadjusted model
```{r}
OUTCOME %>% OR(progression ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(progression ~ marker + age_group + stage_group + fuhrman_group)
```

### Overall mortality
#### Unadjusted model
```{r}
OUTCOME %>% OR(death ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(death ~ marker + age_group + stage_group + fuhrman_group)
```

### Cancer-related mortality
#### Unadjusted model
```{r}
OUTCOME %>% OR(dod ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(dod ~ marker + age_group + stage_group + fuhrman_group)
```

# FOXP3
## Intratumoral lymphocytes
```{r}
OUTCOME <- OUTCOME %>% 
  mutate(marker = foxp3_intratumoral_level)
```

### Tumor progression
#### Unadjusted model
```{r}
OUTCOME %>% OR(progression ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(progression ~ marker + age_group + stage_group + fuhrman_group)
```

### Overall mortality
#### Unadjusted model
```{r}
OUTCOME %>% OR(death ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(death ~ marker + age_group + stage_group + fuhrman_group)
```

### Cancer-related mortality
#### Unadjusted model
```{r}
OUTCOME %>% OR(dod ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(dod ~ marker + age_group + stage_group + fuhrman_group)
```

## Peritumoral lymphocytes
```{r}
OUTCOME <- OUTCOME %>% 
  mutate(marker = foxp3_peritumoral_level)
```

### Tumor progression
#### Unadjusted model
```{r}
OUTCOME %>% OR(progression ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(progression ~ marker + age_group + stage_group + fuhrman_group)
```

### Overall mortality
#### Unadjusted model
```{r}
OUTCOME %>% OR(death ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(death ~ marker + age_group + stage_group + fuhrman_group)
```

### Cancer-related mortality
#### Unadjusted model
```{r}
OUTCOME %>% OR(dod ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(dod ~ marker + age_group + stage_group + fuhrman_group)
```

# CD8
## Intratumoral lymphocytes
```{r}
OUTCOME <- OUTCOME %>% 
  mutate(marker = cd8_intratumoral_level)
```

### Tumor progression
#### Unadjusted model
```{r}
OUTCOME %>% OR(progression ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(progression ~ marker + age_group + stage_group + fuhrman_group)
```

### Overall mortality
#### Unadjusted model
```{r}
OUTCOME %>% OR(death ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(death ~ marker + age_group + stage_group + fuhrman_group)
```

### Cancer-related mortality
#### Unadjusted model
```{r}
OUTCOME %>% OR(dod ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(dod ~ marker + age_group + stage_group + fuhrman_group)
```

## Peritumoral lymphocytes
```{r}
OUTCOME <- OUTCOME %>% 
  mutate(marker = cd8_peritumoral_level)
```

### Tumor progression
#### Unadjusted model
```{r}
OUTCOME %>% OR(progression ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(progression ~ marker + age_group + stage_group + fuhrman_group)
```

### Overall mortality
#### Unadjusted model
```{r}
OUTCOME %>% OR(death ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(death ~ marker + age_group + stage_group + fuhrman_group)
```

### Cancer-related mortality
#### Unadjusted model
```{r}
OUTCOME %>% OR(dod ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(dod ~ marker + age_group + stage_group + fuhrman_group)
```

# CD8-Ki67
## Intratumoral lymphocytes
```{r}
OUTCOME <- OUTCOME %>% 
  mutate(marker = cd8ki67_intratumoral_level)
```

### Tumor progression
#### Unadjusted model
```{r}
OUTCOME %>% OR(progression ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(progression ~ marker + age_group + stage_group + fuhrman_group)
```

### Overall mortality
#### Unadjusted model
```{r}
OUTCOME %>% OR(death ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(death ~ marker + age_group + stage_group + fuhrman_group)
```

### Cancer-related mortality
#### Unadjusted model
```{r}
OUTCOME %>% OR(dod ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(dod ~ marker + age_group + stage_group + fuhrman_group)
```

## Peritumoral lymphocytes
```{r}
OUTCOME <- OUTCOME %>% 
  mutate(marker = cd8ki67_peritumoral_level)

# Note: Not evaluable due to small number of measurements corresponding all to high expression values
table(OUTCOME$marker, useNA = "ifany")
```

### Tumor progression
#### Unadjusted model
```{r}
OUTCOME %>% OR(progression ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(progression ~ marker + age_group + stage_group + fuhrman_group)
```

### Overall mortality
#### Unadjusted model
```{r}
OUTCOME %>% OR(death ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(death ~ marker + age_group + stage_group + fuhrman_group)
```

### Cancer-related mortality
#### Unadjusted model
```{r}
OUTCOME %>% OR(dod ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(dod ~ marker + age_group + stage_group + fuhrman_group)
```

# Ki67
## Intratumoral lymphocytes
```{r}
OUTCOME <- OUTCOME %>% 
  mutate(marker = ki67_intratumoral_level)
```

### Tumor progression
#### Unadjusted model
```{r}
OUTCOME %>% OR(progression ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(progression ~ marker + age_group + stage_group + fuhrman_group)
```

### Overall mortality
#### Unadjusted model
```{r}
OUTCOME %>% OR(death ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(death ~ marker + age_group + stage_group + fuhrman_group)
```

### Cancer-related mortality
#### Unadjusted model
```{r}
OUTCOME %>% OR(dod ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(dod ~ marker + age_group + stage_group + fuhrman_group)
```

# CD8/FOXP3 ratio
## Intratumoral lymphocytes
```{r}
OUTCOME <- OUTCOME %>% 
  mutate(marker = cd8foxp3_intratumoral)
```

### Tumor progression
#### Unadjusted model
```{r}
OUTCOME %>% OR(progression ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(progression ~ marker + age_group + stage_group + fuhrman_group)
```

### Overall mortality
#### Unadjusted model
```{r}
OUTCOME %>% OR(death ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(death ~ marker + age_group + stage_group + fuhrman_group)
```

### Cancer-related mortality
#### Unadjusted model
```{r}
OUTCOME %>% OR(dod ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(dod ~ marker + age_group + stage_group + fuhrman_group)
```

## Peritumoral lymphocytes
```{r}
OUTCOME <- OUTCOME %>% 
  mutate(marker = cd8foxp3_peritumoral)
```

### Tumor progression
#### Unadjusted model
```{r}
OUTCOME %>% OR(progression ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(progression ~ marker + age_group + stage_group + fuhrman_group)
```

### Overall mortality
#### Unadjusted model
```{r}
OUTCOME %>% OR(death ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(death ~ marker + age_group + stage_group + fuhrman_group)
```

### Cancer-related mortality
#### Unadjusted model
```{r}
OUTCOME %>% OR(dod ~ marker)
```

#### Adjusted model: age, pT stage and Fuhrman grade
```{r}
OUTCOME %>% OR(dod ~ marker + age_group + stage_group + fuhrman_group)
```
