---
jupyter:
  jupytext:
    formats: ipynb,Rmd
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.1'
      jupytext_version: 1.1.1
  kernelspec:
    display_name: R
    language: R
    name: ir
---

```{r}
# Library
suppressPackageStartupMessages(library(tidyverse))
library(survival)
library(ggfortify)

# Global
options(warn = -1)

# Scripts
source("Summarize.R")

# Data
load("DF.Rdata")
DF <- DF %>% 
  select(
    death:progression,
    follow_up,
    cd8foxp3_intratumoral:ki67_intratumoral_level
  )

glimpse(DF)
```

# Hazard ratios

```{r}
# A function to estimate hazard ratios using Cox proportional hazards regression
HR_tbl <- function(data, time, event, covariate, str_marker, str_outcome) {
    # Creating the dataset
    time = enquo(time)
    event = enquo(event)
    covariate = enquo(covariate)
    df <- data %>% 
        select(
            time = !! time,
            event = !! event,
            covariate = !! covariate
        ) %>% 
        mutate(event = as.numeric(event))
    # Creating the model
    surv_obj <- survival::Surv(df$time, df$event)
    mdl <- coxph(surv_obj ~ covariate, data = df)
    tbl <- broom::tidy(mdl) %>% 
        mutate(
            hr = round(exp(estimate), 2),
            conf.low = round(exp(conf.low), 2),
            conf.high = round(exp(conf.high), 2),
            p.value = formatC(p.value, digits = 2, format = "g", flag = "#"),
            marker = paste(str_marker),
            outcome = paste(str_outcome)
        )  %>% 
        select(
            outcome, marker, hr, conf.low, conf.high, p.value
        )
    
    return(tbl)
}
```

## Tumor progression

```{r}
str_outcome <- "Tumor progression"
tbl_1A <- DF %>% HR_tbl(
  follow_up, progression, pdl1_tumor_level,
  "PD-L1, tumor cells", str_outcome
)
tbl_1B <- DF %>% HR_tbl(
  follow_up, progression, pdl1_intratumoral_level,
  "PD-L1, intratumoral lymphocytes", str_outcome
)
tbl_1C <- DF %>% HR_tbl(
  follow_up, progression, foxp3_intratumoral_level,
  "FOXP3, intratumoral lymphocytes", str_outcome
)
tbl_1D <- DF %>% HR_tbl(
  follow_up, progression, foxp3_peritumoral_level,
  "FOXP3, peritumoral lymphocytes", str_outcome
)
tbl_1E <- DF %>% HR_tbl(
  follow_up, progression, cd8_intratumoral_level,
  "CD8, intratumoral lymphocytes", str_outcome
)
tbl_1F <- DF %>% HR_tbl(
  follow_up, progression, cd8_peritumoral_level,
  "CD8, peritumoral lymphocytes", str_outcome
)
tbl_1G <- DF %>% HR_tbl(
  follow_up, progression, cd8ki67_intratumoral_level,
  "CD8-Ki67, intratumoral lymphocytes", str_outcome
)
tbl_1H <- DF %>% HR_tbl(
  follow_up, progression, cd8ki67_peritumoral_level,
  "CD8-Ki67, peritumoral lymphocytes", str_outcome
)
tbl_1I <- DF %>% HR_tbl(
  follow_up, progression, ki67_intratumoral_level,
  "Ki67, intratumoral lymphocytes", str_outcome
)
tbl_1J <- DF %>% HR_tbl(
  follow_up, progression, cd8foxp3_intratumoral,
  "CD8/FOXP3, intratumoral lymphocytes", str_outcome
)
tbl_1K <- DF %>% HR_tbl(
  follow_up, progression, cd8foxp3_peritumoral,
  "CD8/FOXP3, peritumoral lymphocytes", str_outcome
)

#  tbl_1H excluded due to NAs
tbl_1 <- bind_rows(
  tbl_1A, tbl_1B,
  tbl_1C, tbl_1D,
  tbl_1E, tbl_1F,
  tbl_1G,
  tbl_1I,
  tbl_1J, tbl_1K
)  %>% 
    as.tibble()

print(tbl_1)
```

## Overall mortality

```{r}
str_outcome <- "Overall mortality"
tbl_2A <- DF %>% HR_tbl(
  follow_up, death, pdl1_tumor_level,
  "PD-L1, tumor cells", str_outcome
)
tbl_2B <- DF %>% HR_tbl(
  follow_up, death, pdl1_intratumoral_level,
  "PD-L1, intratumoral lymphocytes", str_outcome
)
tbl_2C <- DF %>% HR_tbl(
  follow_up, death, foxp3_intratumoral_level,
  "FOXP3, intratumoral lymphocytes", str_outcome
)
tbl_2D <- DF %>% HR_tbl(
  follow_up, death, foxp3_peritumoral_level,
  "FOXP3, peritumoral lymphocytes", str_outcome
)
tbl_2E <- DF %>% HR_tbl(
  follow_up, death, cd8_intratumoral_level,
  "CD8, intratumoral lymphocytes", str_outcome
)
tbl_2F <- DF %>% HR_tbl(
  follow_up, death, cd8_peritumoral_level,
  "CD8, peritumoral lymphocytes", str_outcome
)
tbl_2G <- DF %>% HR_tbl(
  follow_up, death, cd8ki67_intratumoral_level,
  "CD8-Ki67, intratumoral lymphocytes", str_outcome
)
tbl_2H <- DF %>% HR_tbl(
  follow_up, death, cd8ki67_peritumoral_level,
  "CD8-Ki67, peritumoral lymphocytes", str_outcome
)
tbl_2I <- DF %>% HR_tbl(
  follow_up, death, ki67_intratumoral_level,
  "Ki67, intratumoral lymphocytes", str_outcome
)
tbl_2J <- DF %>% HR_tbl(
  follow_up, death, cd8foxp3_intratumoral,
  "CD8/FOXP3, intratumoral lymphocytes", str_outcome
)
tbl_2K <- DF %>% HR_tbl(
  follow_up, death, cd8foxp3_peritumoral,
  "CD8/FOXP3, peritumoral lymphocytes", str_outcome
)

#  tbl_2H excluded due to NAs
tbl_2 <- bind_rows(
  tbl_2A, tbl_2B,
  tbl_2C, tbl_2D,
  tbl_2E, tbl_2F,
  tbl_2G,
  tbl_2I,
  tbl_2J, tbl_2K
)  %>% 
    as.tibble()

print(tbl_2)
```

## Cancer-related mortality

```{r}
str_outcome <- "Cancer-related mortality"
tbl_3A <- DF %>% HR_tbl(
  follow_up, dod, pdl1_tumor_level,
  "PD-L1, tumor cells", str_outcome
)
tbl_3B <- DF %>% HR_tbl(
  follow_up, dod, pdl1_intratumoral_level,
  "PD-L1, intratumoral lymphocytes", str_outcome
)
tbl_3C <- DF %>% HR_tbl(
  follow_up, dod, foxp3_intratumoral_level,
  "FOXP3, intratumoral lymphocytes", str_outcome
)
tbl_3D <- DF %>% HR_tbl(
  follow_up, dod, foxp3_peritumoral_level,
  "FOXP3, peritumoral lymphocytes", str_outcome
)
tbl_3E <- DF %>% HR_tbl(
  follow_up, dod, cd8_intratumoral_level,
  "CD8, intratumoral lymphocytes", str_outcome
)
tbl_3F <- DF %>% HR_tbl(
  follow_up, dod, cd8_peritumoral_level,
  "CD8, peritumoral lymphocytes", str_outcome
)
tbl_3G <- DF %>% HR_tbl(
  follow_up, dod, cd8ki67_intratumoral_level,
  "CD8-Ki67, intratumoral lymphocytes", str_outcome
)
tbl_3H <- DF %>% HR_tbl(
  follow_up, dod, cd8ki67_peritumoral_level,
  "CD8-Ki67, peritumoral lymphocytes", str_outcome
)
tbl_3I <- DF %>% HR_tbl(
  follow_up, dod, ki67_intratumoral_level,
  "Ki67, intratumoral lymphocytes", str_outcome
)
tbl_3J <- DF %>% HR_tbl(
  follow_up, dod, cd8foxp3_intratumoral,
  "CD8/FOXP3, intratumoral lymphocytes", str_outcome
)
tbl_3K <- DF %>% HR_tbl(
  follow_up, dod, cd8foxp3_peritumoral,
  "CD8/FOXP3, peritumoral lymphocytes", str_outcome
)

#  tbl_3H excluded due to NAs
tbl_3 <- bind_rows(
  tbl_3A, tbl_3B,
  tbl_3C, tbl_3D,
  tbl_3E, tbl_3F,
  tbl_3G,
  tbl_3I,
  tbl_3J, tbl_3K
)  %>% 
    as.tibble()

print(tbl_3)
```

## All outcomes

**Forest plots of biomarkers by outcome.** Hazards ratios (dots) with 95% confidence intervals (error bars) for tumor progression, overall mortality and cancer-related mortality by expression levels of PD-L1, FOXP3, CD8, CD8-Ki67 and CD8/FOXP3 (high level vs low level). The dashed line corresponds to a hazard ratio of 1. Hazards ratios (HR) with their 95% confidence intervals (minimum, maximum) and P values were estimated using Cox's proportional hazards regression models.
```{r}
options(repr.plot.height = 9, repr.plot.width = 9)

tbl_HR <- bind_rows(tbl_1, tbl_2, tbl_3)

Plot_HR <- tbl_HR  %>% 
  mutate(
    outcome = factor(
      outcome,
      levels = c("Tumor progression", "Overall mortality", "Cancer-related mortality")
    ),
    marker = fct_rev(
      factor(marker,
        levels = c(
          "PD-L1, tumor cells",
          "PD-L1, intratumoral lymphocytes",
          "FOXP3, intratumoral lymphocytes",
          "FOXP3, peritumoral lymphocytes",
          "CD8, intratumoral lymphocytes",
          "CD8, peritumoral lymphocytes",
          "CD8-Ki67, intratumoral lymphocytes",
          "Ki67, intratumoral lymphocytes",
          "CD8/FOXP3, intratumoral lymphocytes",
          "CD8/FOXP3, peritumoral lymphocytes"
      )))
    ) %>% 
    ggplot(aes(x = marker, y = hr, ymin = conf.low, ymax = conf.high)) +
      geom_pointrange() +
      geom_errorbar(aes(ymin = conf.low, ymax = conf.high)) +
      geom_hline(yintercept = 1, linetype = 2) +
      labs(x = NULL, y = "Hazard ratio") +
      scale_y_continuous(limits = c(0, 11), breaks = c(0:11)) +
      coord_flip(clip = "off") +
      facet_wrap(~ outcome, ncol = 1) +
      geom_text(
        aes(
          label = paste0(" HR=", hr, " (", conf.low, ", ", conf.high, "), P=", p.value),
            y = Inf,
            x = marker
          ), hjust = 0, size = 3
      ) +
      theme(plot.margin = unit(c(1,16,1,1), "lines"))

Plot_HR
```

```{r}
ggsave(
    filename = "Figs/PFCK-papRCC-HR.png",
    Plot_HR,
    width = 9,
    height = 9,
    units = "in"
)
```

# Survival curves

```{r}
# A function to plot survival curves with P values from the Mantel-Cox test

Plot_Surv <- function(data, time, event, covariate) {
    # Data
    time = enquo(time)
    event = enquo(event)
    covariate = enquo(covariate)
    df <- data %>% 
        select(
            time = !! time,
            event = !! event,
            covariate = !! covariate
      )
    
    # logrank test
    diff <- survdiff(Surv(time, as.numeric(event)) ~ covariate, data = df)
    diff_glance <- broom::glance(diff)
    diff_p <- diff_glance$p.value %>% formatC(format = "g", flag = "#", digits = 2)
    
    # Survival plot
    fit <- survfit(Surv(time, as.numeric(event)) ~ covariate, data = df)
    plot <- autoplot(fit, conf.int = FALSE, surv.size = 1, censor = FALSE) +
        labs(x = NULL, y = NULL, caption = paste0("P=", diff_p)) +
        theme(
            legend.justification = c("left", "top"),
            plot.caption = element_text(size = 11),
            legend.title = element_blank()
        )
    
    return(plot)
}
```

## Tumor progression
**Survival curves for tumor progression by biomarkers.** The x-axes show follow-up in months. The y-axes show survival probabilities. P values were estimated using the Mantel-Cox (log-rank) test.

```{r}
options(repr.plot.height = 18, repr.plot.width = 12)

Surv_1A <- DF %>% 
    Plot_Surv(follow_up, progression, pdl1_tumor_level) +
        labs(tag = "A", title = "PD-L1 in tumor cells")
Surv_1B <- DF %>% 
    Plot_Surv(follow_up, progression, pdl1_intratumoral_level) +
        labs(tag = "B", title = "PD-L1 in intratumoral lymphocytes")
Surv_1C <- DF %>% 
    Plot_Surv(follow_up, progression, foxp3_intratumoral_level) +
        labs(tag = "C", title = "FOXP3 in intratumoral lymphocytes")
Surv_1D <- DF %>% 
    Plot_Surv(follow_up, progression, foxp3_peritumoral_level) +
        labs(tag = "D", title = "FOXP3 in peritumoral lymphocytes")
Surv_1E <- DF %>% 
    Plot_Surv(follow_up, progression, cd8_intratumoral_level) +
        labs(tag = "E", title = "CD8 in intratumoral lymphocytes")
Surv_1F <- DF %>% 
    Plot_Surv(follow_up, progression, cd8_peritumoral_level) +
        labs(tag = "F", title = "CD8 in peritumoral lymphocytes")
Surv_1G <- DF %>% 
    Plot_Surv(follow_up, progression, cd8ki67_intratumoral_level) +
        labs(tag = "G", title = "CD8-Ki67 in intratumoral lymphocytes")
Surv_1H <- DF %>% 
    Plot_Surv(follow_up, progression, ki67_intratumoral_level) +
        labs(tag = "H", title = "Ki67 in intratumoral lymphocytes")
Surv_1I <- DF %>% 
    Plot_Surv(follow_up, progression, cd8foxp3_intratumoral) +
        labs(tag = "I", title = "CD8/FOXP3 ratio in intratumoral lymphocytes")
Surv_1J <- DF %>% 
    Plot_Surv(follow_up, progression, cd8foxp3_peritumoral) +
        labs(tag = "J", title = "CD8/FOXP3 ratio in peritumoral lymphocytes")

Surv_1 <- gridExtra::grid.arrange(
    Surv_1A, Surv_1B,
    Surv_1C, Surv_1D,
    Surv_1E, Surv_1F,
    Surv_1G, Surv_1H,
    Surv_1I, Surv_1J,
    ncol = 2
)
```
